---
title: "covid_data_exploration"
author: "Nora Ghenciulescu"
date: "2025-05-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
theme_set(theme_bw())
library(lubridate)
library(segmented)
library(surveillance)
library(MASS)
library(nlme)

library(future)
library(future.apply)
plan("multisession")
options(mc.cores = 6)

library(nleqslv)


# this is an adaptation of the cp_fun.R file of the original paper
# https://zenodo.org/records/4449816
source("code/help_functions.R")
```

This .Rmd file should be run in a directory where it case access to code/help_functions.R as well as to the data folder that contains the age_group_data.csv and the total_age_distribution.csv. This can be done by using the repo created at https://github.com/ReggieScript/InfectiousDisease-COVIDMX/tree/main.


### Overall model

For K = 2,...,6 breakpoints, we run model on the data for the 2020-02-19 - 2020-05-15 period.

Data prep:

```{r}
data <- read_csv("../data/age_group_data.csv", show_col_types = FALSE)

# we only need the 'onsets' column, summed over all age groups; also replace NAs and sort by 'date'
data_full = data %>% group_by(date) %>%
  summarise(onsets=sum(onsets)) %>%
  right_join(tibble(date=seq(ymd("2020-02-19"), ymd("2020-05-15"), by = "1 day"))) %>% 
  arrange(date) %>%
  mutate(onsets=replace_na(onsets,0))

data_full
```

Plot:
```{r}
# we want a plot for the full data, not just the 2020-02-19 - 2020-05-15 period!

data_fullyear = data %>% group_by(date) %>%
  summarise(onsets=sum(onsets)) %>%
  right_join(tibble(date=seq(ymd("2020-02-19"), ymd("2020-12-31"), by = "1 day"))) %>% 
  arrange(date) %>%
  mutate(onsets=replace_na(onsets,0))

# again summarize onsets
summary_onsets <- data_fullyear %>%
  group_by(date) %>%
  summarise(total_onsets = sum(onsets, na.rm = TRUE))

# specify lockdown start and lift dates 
lockdown_dates <- data.frame(
  date = as.Date(c("2020-03-31", "2020-06-01")),
  label = c("Lockdown Start", "Lockdown Lift")
)

# plot
ggplot(summary_onsets, aes(x = date, y = total_onsets)) +
  geom_col(width = 1) +
  geom_vline(data = lockdown_dates, aes(xintercept = date, color = label),
             linetype = "dashed", size = 1) +
  scale_color_manual(name = "Events",
                     values = c("Lockdown Start" = "red", "Lockdown Lift" = "blue")) +
  labs(title = "Daily CPVOD-19 Case Onsets",
       x = "Date",
       y = "Total Onsets") +
  theme_minimal() +
  theme(legend.position = "top")
```

Perform the change point analysis:

```{r}
# this uses the perform_cp_analysis() in the cp_fun.R file, but modified a bit
# this does backprojection, discrete optimisation of breakpoints and fits the segmented regression model
# returns an object which contains the models and the identified break points for all values of K = 2,...,6

cp_res_full = perform_cp_analysis(data = data_full,
                                      type = "both",
                                      cp_max_onset = 6,
                                      cp_max_backpro = 6,
                                      save_disc_optim_results = T,
                                      use_disc_optim_results = T,
                                      name_disc =  "saved_data")
cp_res_full
```

Extract BIC:

```{r}
cp_res_full['bic_backpro']
```

We see that the optimal number of breakpoints is 3 break points!


### Models per age group

```{r}
# split data per age group:
data_014 = data %>% dplyr::filter(age_group=="0-14") %>%
  group_by(date) %>%
  summarise(onsets=sum(onsets)) %>%
  right_join(tibble(date=seq(ymd("2020-02-24"), ymd("2020-05-15"), by = "1 day"))) %>%
  arrange(date) %>%
  mutate(onsets=replace_na(onsets,0))

data_1559 = data %>% dplyr::filter(age_group=="15-59") %>%
  group_by(date) %>%
  summarise(onsets=sum(onsets)) %>%
  right_join(tibble(date=seq(ymd("2020-02-24"), ymd("2020-05-15"), by = "1 day"))) %>%
  arrange(date) %>%
  mutate(onsets=replace_na(onsets,0))

data_6079 = data %>% dplyr::filter(age_group=="60-79") %>%
  group_by(date) %>%
  summarise(onsets=sum(onsets)) %>%
  right_join(tibble(date=seq(ymd("2020-02-24"), ymd("2020-05-15"), by = "1 day"))) %>%
  arrange(date) %>%
  mutate(onsets=replace_na(onsets,0))

data_80 = data %>% dplyr::filter(age_group=="80+") %>%
  group_by(date) %>%
  summarise(onsets=sum(onsets)) %>%
  right_join(tibble(date=seq(ymd("2020-02-24"), ymd("2020-05-15"), by = "1 day"))) %>%
  arrange(date) %>%
  mutate(onsets=replace_na(onsets,0))
```

```{r}
# run model for each age group:
cp_res_014 = perform_cp_analysis(data = data_014, 
                                      type = "backpro",
                                      cp_max_onset = 6, 
                                      cp_max_backpro = 6,
                                      save_disc_optim_results = T, 
                                      use_disc_optim_results = T, 
                                      name_disc =  "saved_data_014")

# select the model with smallest BIC:
res_014_selected_backpro = cp_res_014$cp_segmented_list_backpro[[which.min(cp_res_014$bic_backpro)]]
```

```{r}
cp_res_1559 = perform_cp_analysis(data = data_1559, 
                                      type = "backpro",
                                      cp_max_onset = 6, 
                                      cp_max_backpro = 6,
                                      save_disc_optim_results = T, 
                                      use_disc_optim_results = T, 
                                      name_disc =  "saved_data_1559")

res_1559_selected_backpro = cp_res_1559$cp_segmented_list_backpro[[which.min(cp_res_1559$bic_backpro)]]
```

```{r}
cp_res_6079 = perform_cp_analysis(data = data_6079, 
                                      type = "backpro",
                                      cp_max_onset = 6, 
                                      cp_max_backpro = 6,
                                      save_disc_optim_results = T, 
                                      use_disc_optim_results = T, 
                                      name_disc =  "saved_data_6079")

res_6079_selected_backpro = cp_res_6079$cp_segmented_list_backpro[[which.min(cp_res_6079$bic_backpro)]]
```

```{r}
# note: the 80+ group only works for 2 break points (cp_max_onset = 2, cp_max_backpro = 2), we don't know why
cp_res_80 = perform_cp_analysis(data = data_80, 
                                      type = "backpro",
                                      cp_max_onset = 2, 
                                      cp_max_backpro = 2,
                                      save_disc_optim_results = T, 
                                      use_disc_optim_results = T, 
                                      name_disc =  "saved_data_80")

res_80_selected_backpro = cp_res_80$cp_segmented_list_backpro[[which.min(cp_res_80$bic_backpro)]]
```


```{r}
# load age distribution file
age = read.csv("../data/total_age_distribution.csv")
age_file = age %>% mutate(age_group=cut(AGE, c(-1,14,59,79,120), labels = c("0-14", "15-59", "60-79", "80+"))) %>%
  group_by(age_group) %>%
  summarise(n=sum(num)) 


# plot:
plot_age_group = res_014_selected_backpro$segmented_model$model %>% 
  dplyr::select(t, logbackpro) %>%
  mutate(age_group="0-14") %>%
  cbind(pred=res_014_selected_backpro$segmented_model$fitted) %>%
  rbind(res_1559_selected_backpro$segmented_model$model %>% 
          dplyr::select(t, logbackpro) %>%
          mutate(age_group="15-59") %>%
          cbind(pred=res_1559_selected_backpro$segmented_model$fitted)) %>%
  rbind(res_6079_selected_backpro$segmented_model$model %>% 
          dplyr::select(t, logbackpro) %>%
          mutate(age_group="60-79") %>%
          cbind(pred=res_6079_selected_backpro$segmented_model$fitted)) %>%
  rbind(res_80_selected_backpro$segmented_model$model %>% 
          dplyr::select(t, logbackpro) %>%
          mutate(age_group="80+") %>%
          cbind(pred=res_80_selected_backpro$segmented_model$fitted)) %>%
  mutate(t=ymd("2020-02-27")+t) %>%
  right_join(age_file) %>%
  mutate(pred_per_100k = (exp(pred)/n)*100000) %>%
  ggplot() +
  geom_line(aes(t, pred_per_100k, color=age_group)) +
  ylab("Number infections\n per 100k") + xlab("Date") +
  theme(
    axis.text=element_text(size = rel(1.3)),
    axis.title=element_text(size = rel(1.3)),
    legend.text = element_text(size = rel(1.3)),
    legend.title =element_text(size = rel(1.3)))+
  guides(lty=guide_legend(title="Age group"))

plot_age_group + ggtitle("Age group") + theme()
```